Cognate beginnings to bilingual lexical acquisition

### 2023-08-13 manuscript/_appendix/_model-diagnostics.qmd
Fix IC's email, add again the model diagnostics appendix, and add notes in formula

```{r fig-rhats-neffs}
#| label: fig-rhats-neffs
#| fig-cap: "MCMC convergence diagnostic of all parameters in the model. A: distribution of the Gelman-Rubin (R-hat) scores. B: distribution of the ratio of effective sa"
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3
#| fig-width: 6
#| out-width: 100%
model_convergence |> 
    ggplot(aes(.rhat)) +
    geom_histogram(binwidth = 0.001, colour = "white", fill = "grey60") +
    geom_vline(xintercept = 1.01, linetype = "dashed") +
    labs(x = "R-hat",
         y = "MCMC samples") +

    model_convergence |> 
    ggplot(aes(.neff)) +
    geom_histogram(binwidth = 0.1, colour = "white", fill = "grey60") +
    geom_vline(xintercept = 1.01, linetype = "dashed") +
    labs(x = "Effective sample size ratio",
         y = "MCMC samples") +

    plot_layout(ncol = 1) &
    plot_annotation(tag_levels = "A") &
    theme(legend.position = "none",
          panel.grid = element_blank())
```

@@

```{r}
#| label: fig-model-pairs
#| fig-cap: "Marginal distribution and bi-variate scatterplot of posterior samples for the fixed regression coefficients in Model 3."
#| fig-height: 8
#| fig-width: 10
#| out-width: 100%
#| out-height: 100%
#| echo: false
#| warning: false
#| message: false
titles <- c("Comprehension",
            "Production",
            "Age",
            "Length",
            "Exposure",
            "Cognateness",
            "Age × Exposure",
            "Age × Cognateness",
            "Exposure × Cognateness",
            "Age × Exposure × Cognateness")

@@

my_diag <- function(data, mapping, ...) {
    ggplot(data = data, mapping = mapping) + 
        stat_slab(colour = "white",
                  fill = "grey15")
}

my_upper <- function(data, mapping, ...) {
    ggplot(data = data, mapping = mapping) + 
        stat_density_2d(aes(fill = ..density..),
                        geom = "raster",
                        contour = FALSE) +
        scale_fill_distiller(palette = "Greys",
                             direction = 1) 
}

posterior %>%
    GGally::ggpairs(
        upper = list(continuous = my_upper),
        diag = list(continuous = my_diag),
        lower = NULL,
        columnLabels = c("Intercept\n(Comprehension)",
                         "Intercept\n(Production)",
                         "Age", "Length", "Exposure",
                         "Levenshtein", "Age × Exposure",
                         "Age × Levenshtein", "Exposure × Leveshtein",
                         "Age × Exposure ×\nLevenshtein"),
        axisLabels = "none"
    ) +
    theme(strip.text = element_text(size = 5))

@@

```{r tbl-ppc}
#| label: fig-ppc
#| fig-cap: "Model posterior predictive checks (PPC). Bars indicate the observed proportion of responses to each category (No, Understands, and Understands and Says). Blue dots and error bars represent the mean proportion of responses simulated from the posterior for each category, and its 95\\% interval."
#| eval: true
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6
#| fig-height: 3
#| out-width: 80%
#| out-height: 80%
y_int <- as.integer(as.integer(responses$response))
bayesplot::color_scheme_set(c(rep("grey60", 2), rep("black", 4)))

bayesplot::ppc_bars(
    y_int, 
    model_ppcs,
    prob = 0.95,
    linewidth = 1,
    size = 1.25,
    fatten = 2,
    freq = FALSE
) +
    scale_x_continuous(breaks = 1:3,
                       labels = c("No",
                                  "Understands", 
                                  "Understands and Says")) +
    theme(legend.position = "none",
          axis.text = element_text(size = 11))
```

>>>>

```{r fig-rhats-neffs}
#| label: fig-rhats-neffs
#| fig-cap: "MCMC convergence diagnostic of all parameters in the model. Each dot represents the score of one parameter. (A) Distribution of the Gelman-Rubin (R-hat) scores. (B) Distribution of the ratio of effective sample size."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3
#| fig-width: 6
#| out-width: 100%
model_convergence |> 
    ggplot(aes(.rhat)) +
    geom_dots(colour = "black") +
    geom_vline(xintercept = 1.01, linetype = "dashed") +
    labs(
        x = "R-hat",
        y = "MCMC samples"
    ) +
    model_convergence |> 
    ggplot(aes(.neff)) +
    geom_dots(colour = "black") +
    geom_vline(xintercept = 1.01, linetype = "dashed") +
    labs(
        x = "Effective sample size ratio",
        y = "MCMC samples"
    ) +    
    plot_layout(ncol = 1) &
    plot_annotation(tag_levels = "A") &
    theme(
        legend.position = "none",
        panel.grid = element_blank()
    )
```

@@

```{r}
#| label: fig-model-pairs
#| fig-cap: "Marginal distribution and bi-variate scatterplot of posterior samples for the fixed regression coefficients in Model 3."
#| fig-height: 8
#| fig-width: 10
#| out-width: 100%
#| out-height: 100%
#| echo: false
#| warning: false
#| message: false
titles <- c(
    "Comprehension",
    "Production",
    "Age",
    "Length",
    "Exposure",
    "Cognateness",
    "Age × Exposure",
    "Age × Cognateness",
    "Exposure × Cognateness",
    "Age × Exposure × Cognateness"
)

@@

my_diag <- function(data, mapping, ...) {
    ggplot(data = data, mapping = mapping) +
        stat_slab(
            colour = "white",
            fill = "grey15"
        )
}

my_upper <- function(data, mapping, ...) {
    ggplot(data = data, mapping = mapping) +
        stat_density_2d(aes(fill = ..density..),
            geom = "raster",
            contour = FALSE
        ) +
        scale_fill_distiller(
            palette = "Greys",
            direction = 1
        )
}

posterior |> 
    GGally::ggpairs(
        upper = list(continuous = my_upper),
        diag = list(continuous = my_diag),
        lower = NULL,
        columnLabels = c(
            "Intercept\n(Comprehension)",
            "Intercept\n(Production)",
            "Age", "Length", "Exposure",
            "Levenshtein", "Age × Exposure",
            "Age × Levenshtein", "Exposure × Leveshtein",
            "Age × Exposure ×\nLevenshtein"
        ),
        axisLabels = "none"
    ) +
    theme(
        strip.text = element_text(size = 5),
        panel.grid = element_blank(),
        panel.border = element_blank()
    )

@@

```{r tbl-ppc}
#| label: fig-ppc
#| fig-cap: "Model posterior predictive checks (PPC). Bars indicate the observed proportion of responses to each category (No, Understands, and Understands and Says). Blue dots and error bars represent the mean proportion of responses simulated from the posterior for each category, and its 95\\% interval."
#| eval: true
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6
#| fig-height: 3
#| out-width: 80%
#| out-height: 80%
y_int <- as.integer(as.integer(model_fit$data$response))
bayesplot::color_scheme_set(c(rep("grey60", 2), rep("black", 4)))

bayesplot::ppc_bars(
    y_int,
    model_ppcs,
    prob = 0.95,
    linewidth = 1,
    size = 1.25,
    fatten = 2,
    freq = FALSE
) +
    scale_x_continuous(
        breaks = 1:3,
        labels = c(
            "No",
            "Understands",
            "Understands and Says"
        )
    ) +
    theme(
        legend.position = "none",
        axis.text = element_text(size = 11),
        panel.grid = element_blank()
    )
```

### 2023-08-13 manuscript/appendix.qmd
Fix IC's email, add again the model diagnostics appendix, and add notes in formula

```{r setup}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load objects -----------------------------------------------------------------
library(knitr)
library(kableExtra)
library(ggplot2)

targets::tar_config_set(store = here::here("_targets"),
                        script = here::here("_targets.R"))

@@

targets::tar_load(c(bvq_data, childes, items, model_summary,
                    participants, responses))

@@

## Appendix A: frequency and language exposure as separate predictors {.appendix .unnumbered}

@@

>>>>

```{r setup}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load objects -----------------------------------------------------------------
library(knitr)
library(kableExtra)
library(ggplot2)
library(ggdist)
library(patchwork)

targets::tar_config_set(
    store = here::here("_targets"),
    script = here::here("_targets.R")
)

@@

targets::tar_load(c(
    bvq_data, childes, items, model_summary,
    participants, responses
))

targets::tar_load(c(model_fit, model_convergence, model_ppcs))

@@

## Appendix A: model diagnostics {.appendix .unnumbered}

{{< include _appendix/_model-diagnostics.qmd >}}

@@

## Appendix C: frequency and language exposure as separate predictors {.appendix .unnumbered}

### 2023-08-13 manuscript/manuscript.qmd
Fix IC's email, add again the model diagnostics appendix, and add notes in formula

---
abstract: |
    Bilingual infants' developmental trajectories of lexical acquisition are equivalent to their monolingual peers'. This is remarkable, given the complexity of their linguistic input. Recent studies suggest that bilingual vocabulary growth is boosted by the number of cognates (form-similar translation equivalents) shared by the pair of languages being learned, and that this cognateness facilitation effect is driven by a stronger parallel activation of cognates during linguistic exposure, compared to non-cognates. The mechanisms behind this facilitation are still unclear. In this study, we propose an account of bilingual lexical acquisition in which parallel activation increases the rate at which children accumulate learning instances for words in both languages, even in fully monolingual situations. We predicted a stronger cognate facilitation for words to which children were exposed less frequently (low-exposure words), as they are co-activated by their translation more often than high-exposure words. We developed an extensive online vocabulary checklist, the Barcelona Vocabulary questionnaire (BVQ), to collect vocabulary data from 366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We used Bayesian explanatory item response theory to model the acquisition trajectories of 604 Catalan and Spanish words. We found an interaction between exposure and cognateness, suggesting that cognateness facilitates the aquisition of low-exposure words, but not of mean exposure or high-exposure words. Overall, our findings suggest that cognateness plays a key role in bilingual lexical acquisition, and provides evidence for a frequency-mediated facilitation effect driven by parallel activation.

@@

$$
\begin{aligned}
&\textbf{Response}~(k)~\textbf{to word}~i~ \textbf{by participant}~j \\
\text{Response}_{ij} &\sim \text{Cumulative logit}(\theta_{k_{ij}}) \\ 
&\text{where}~k \in \{\text{No} \rightarrow \text{Understands}, \text{Understands} \rightarrow \text{Understands and Says}\}\\
&\textbf{Distribution parameters} \\
\theta_{k_{ij}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + (\beta_{1} + u_{1_{i}} + w_{1_{j}}) \cdot \text{Age}_{i} + \\
&(\beta_{2} + u_{2_{i}} + w_{2_{j}}) \cdot \text{Length}_{ij} + 
(\beta_{3} + u_{3_{i}} + w_{3_{j}}) \cdot \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) \cdot \text{Cognateness}_{ij} + (\beta_{5} + u_{5_{i}} + w_{3_{j}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) \cdot (\text{Age}_{i} \times \text{Cognateness}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) \cdot (\text{Exposure}_{ij} \times \text{Cognateness}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Cognateness}_{ij}) \\
&\text{where:}\\
& u_{1-8_{i}}~\text{: participant-level adjustments} \\
& w_{1-3_{j}}~\text{: TE-level adjustments} \\
&\textbf{Prior} \\
\beta_{0_{k}} &\sim \mathcal{N}(-0.25, 0.5); ~\beta_{1-5} \sim \mathcal{N}(0, 1) \\
\sigma_{u_{0-8}, w_{0-3}} &\sim \mathcal{N_{+}}(1, 0.25); ~\rho_{u_{0-8}, w_{0-3}} \sim \text{LKJcorr}(2) \\
&\text{where }~\rho_{u_{0-8}, w_{0-3}}~\text{are the correlations between group-level adjustments}\\
\end{aligned}
$$ {#eq-model}

@@

Chain convergence diagnostics showed adequate values, and the model posterior showed little evidence of correlation, as indicated by negligible pairwise correlation between the marginal posterior distribution of coefficients. @tbl-coefs shows the summary of the posterior distribution of the fixed regression coefficients, and their degree of overlap with the ROPE. For interpretability, we report the estimated regression coefficients transformed to the probability scale^[The logit and probability scales relate non-linearly. This means that one logit difference is not necessarily translated to a unique value in the probability scale. For example, the probability of acquisition of a given word might increase in 5% when age increases from 22 to 23 months, the probability of acquisition of the same word might only increase in 0.2% when age increases from 31 to 32 months. The linear growth of the probability of acquisition differs along the logistic curve, and therefore deciding the age point at which to report the estimates of the regression coefficients in the probability scale is not trivial. Following @gelman2020regression, we report the maximum value of such coefficient, which corresponds to the linear growth (i.e. derivative) of the logistic curve at the age at which most participants were acquiring a given word. This value can be approximated by dividing the coefficient in the logit scale by four: $\hat{\beta_j}/4$, where $\hat{\beta_j}$ is the estimated mean of the posterior distribution of coefficient $j$.]. The resulting values correspond to the maximum difference in probability of acquisition (*Comprehension* or *Comprehension and Production*) that corresponds to a one standard deviation change in each predictor.

@@

An additional analysis including lexical frequency and language exposure as separate predictors (instead of the composite *Exposure* measure) showed equivalent results (see Appendix A). To rule out the possibility that cognateness facilitation effect we found was due to cognateness comprising more frequent syllables than non-cognates---and therefore not because of their cognate status itself---, we compared the syllabic frequency of cognates and non-cognates included in our analyses. To calculate syllable frequency, we first extracted all syllables embedded in the selected words. For each syllable, we summed the lexical frequency of all the words in which such syllable appeared. The resulting value provided an estimate of the number of times the syllable appears in child-directed speech, embedded within different words. Finally, for each word-form, we summed the frequency of its syllables, as an estimate of the syllabic frequency of the word-form. We fit a Bayesian model with *Cognateness* as response variable, and the main effects of syllable frequency and number of syllables (to control for the fact words with more syllables are more likely to score higher in syllabic frequency) as predictors. This model provided strong evidence for the association between cognateness and syllabic frequency being equivalent to zero (see Appendix B).

>>>>

---
abstract: |
    Bilingual infants' developmental trajectories of lexical acquisition are equivalent to their monolingual peers'. This is remarkable, given the complexity of their linguistic input. Recent studies suggest that bilingual vocabulary growth is boosted by the number of cognates (form-similar translation equivalents) shared by the pair of languages being learned, and that this cognateness facilitation effect is driven by a stronger parallel activation of cognates during linguistic exposure, compared to non-cognates. The mechanisms behind this facilitation are still unclear. In this study, we propose an account of bilingual lexical acquisition in which parallel activation increases the rate at which children accumulate learning instances for words in both languages, even in fully monolingual situations. We predicted a stronger cognate facilitation for words to which children were exposed less frequently (low-exposure words), as they are co-activated by their translation more often than high-exposure words. We developed an extensive online vocabulary checklist, the Barcelona Vocabulary questionnaire (BVQ), to collect vocabulary data from 366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We used Bayesian explanatory item response theory to model the acquisition trajectories of 604 Catalan and Spanish words. We found an interaction between exposure and cognateness, suggesting that cognateness facilitates the aquisition of low-exposure words, but not of mean exposure or high-exposure words. Overall, our findings suggest that cognateness plays a key role in bilingual lexical acquisition, and provide evidence for a frequency-mediated facilitation effect driven by parallel activation.

@@

$$
\begin{aligned}
&\textbf{Response}~(k)~\textbf{to word}~i~ \textbf{by participant}~j \\
\text{Response}_{ij} &\sim \text{Cumulative logit}(\theta_{k_{ij}}) \\ 
&\text{where}~k \in \{\text{No} \rightarrow \text{Understands}, \text{Understands} \rightarrow \text{Understands and Says}\}\\
&\textbf{Distribution parameters} \\
\theta_{k_{ij}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + (\beta_{1} + u_{1_{i}} + w_{1_{j}}) \cdot \text{Age}_{i} + \\
&(\beta_{2} + u_{2_{i}} + w_{2_{j}}) \cdot \text{Length}_{ij} + 
(\beta_{3} + u_{3_{i}} + w_{3_{j}}) \cdot \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) \cdot \text{Cognateness}_{ij} + (\beta_{5} + u_{5_{i}} + w_{3_{j}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) \cdot (\text{Age}_{i} \times \text{Cognateness}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) \cdot (\text{Exposure}_{ij} \times \text{Cognateness}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Cognateness}_{ij}) \\
&\text{where:}\\
& \beta_{1-8}\text{: fixed effects}\\
& u_{1-8_{i}}\text{: participant-level random effects} \\
& w_{1-3_{j}}\text{: TE-level random effects} \\
&\textbf{Prior} \\
\beta_{0_{k}} &\sim \mathcal{N}(-0.25, 0.5); ~\beta_{1-5} \sim \mathcal{N}(0, 1) \\
\sigma_{u_{0-8}, w_{0-3}} &\sim \mathcal{N_{+}}(1, 0.25); ~\rho_{u_{0-8}, w_{0-3}} \sim \text{LKJcorr}(2) \\
&\text{where:}\\
& \sigma_{u_{0-8}, w_{0-3}}\text{: are the standard deviations of }u\text{ and }w\\
& \rho_{u_{0-8}, w_{0-3}}~\text{are the correlations between random effects in }u\text{ and }w\\
\end{aligned}
$$ {#eq-model}

@@

The model posterior showed adequate chain convergence diagnostics and posterior predictive checks (see Appendix A). @tbl-coefs shows the summary of the posterior distribution of the fixed regression coefficients, and their degree of overlap with the ROPE. For interpretability, we report the estimated regression coefficients transformed to the probability scale^[The logit and probability scales relate non-linearly. This means that one logit difference is not necessarily translated to a unique value in the probability scale. For example, the probability of acquisition of a given word might increase in 5% when age increases from 22 to 23 months, the probability of acquisition of the same word might only increase in 0.2% when age increases from 31 to 32 months. The linear growth of the probability of acquisition differs along the logistic curve, and therefore deciding the age point at which to report the estimates of the regression coefficients in the probability scale is not trivial. Following @gelman2020regression, we report the maximum value of such coefficient, which corresponds to the linear growth (i.e. derivative) of the logistic curve at the age at which most participants were acquiring a given word. This value can be approximated by dividing the coefficient in the logit scale by four: $\hat{\beta_j}/4$, where $\hat{\beta_j}$ is the estimated mean of the posterior distribution of coefficient $j$.]. The resulting values correspond to the maximum difference in probability of acquisition (*Comprehension* or *Comprehension and Production*) that corresponds to a one standard deviation change in each predictor.

@@

An additional analysis including lexical frequency and language exposure as separate predictors (instead of the composite *Exposure* measure) showed equivalent results (see Appendix B). To rule out the possibility that cognateness facilitation effect we found was due to cognateness comprising more frequent syllables than non-cognates---and therefore not because of their cognate status itself---, we compared the syllabic frequency of cognates and non-cognates included in our analyses. To calculate syllable frequency, we first extracted all syllables embedded in the selected words. For each syllable, we summed the lexical frequency of all the words in which such syllable appeared. The resulting value provided an estimate of the number of times the syllable appears in child-directed speech, embedded within different words. Finally, for each word-form, we summed the frequency of its syllables, as an estimate of the syllabic frequency of the word-form. We fit a Bayesian model with *Cognateness* as response variable, and the main effects of syllable frequency and number of syllables (to control for the fact words with more syllables are more likely to score higher in syllabic frequency) as predictors. This model provided strong evidence for the association between cognateness and syllabic frequency being equivalent to zero (see Appendix C).
